---
title: "Data Frames and Visualisation"
author: "Katy Brown"
date: "2026-07-03"
output:
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/MB1A_Practicals")
library(tidyverse)
library(gt)
library(maps)
library(learnr)
library(knitr)
```

## Data Frames and Visualisation - Part 1

### Objectives

In this practical we will learn to:

* Load dataframes into R
* Calculate dataframe summary statistics
* Select specific columns and rows
* Merge data from multiple dataframes
* Convert data from "wide" to "long" format
* Visualise data using scatter plots

## Dataset
For this worked example, we will use data from [this Science paper](https://doi.org/10.1126/science.ads0532), published in June 2026.

This study tracked the migration of a songbird species, pied flycatchers (*Ficedula hypoleuca*). We will reproduce Figure 1B from this paper, which shows the breeding site, Iberian stopover location and winter location for 8 bird populations. All of the data we will use comes from the Supplementary Data associated with this publication.


## CSV Format
The data we will use in this practical is stored in comma-separated values (CSV) format, a plain-text file format where each line represents a row and values in columns are seperated by commas.

For example, if we have the following table:


```{r csv_example, echo=FALSE}
df <- data.frame(
          fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(df)
```

The contents of the CSV file would be:
```{text}
fruit,length,weight
apple,22,150
pear,30,165
banana,55,170
```

### Practical Data

The first part of the dataset we will use for this practical contains the longitude and latitude co-ordinates for breeding sites for eight specific populations of pied flycatcher.

The file is in the data directory and named `breeding_site_locations.csv`. Open this file in a text editor (for example NotePad, TextEdit) and examine it.

```{r question_moscow, echo = FALSE}
question_text(
"
Quiz Question\n
What is the longitude of the Moscow breeding site?",
answer(36.7, correct=TRUE),
allow_retry = TRUE)
```

## Reading a CSV file

Now we will start to build an R script, inside an Rmarkdown document, to examine this dataset.

The R function `read_csv` reads a CSV file into R. 

If I wanted to read the file `weather_data.csv`, the code I would use would be:

```{r read_weather_data}
weather <- read_csv("Dataframes/data/weather_data.csv")
```

And to view the table I would just type the variable name do the following:

```{r show_weather_data}
weather
```

In your Rmarkdown document, read `breeding_site_locations.csv` into a variable, then view it.

```{r read_bird_data, include=FALSE}
bird_breeding_sites <- read_csv("Dataframes/data/breeding_site_locations.csv")
```

It should look like this:

```{r view_breeding_location_data}
bird_breeding_sites
```

## Plotting bird breeding sites

Now we can draw a scatter plot.

```{r initiate_plot_1}
ggplot(bird_breeding_sites,
       aes(
         x = breeding_lon,
         y = breeding_lat,
        )
       ) +
  geom_point() +
  annotation_borders(
    linewidth=0.2
  )
```

Read the table containing the bird migration data.

```{r read_main_dataframe}
bird_migration = read_csv("Dataframes/data/migratory_connectivity_dataset.csv")
```


```{r view_main_dataframe}
gt(bird_migration)
```

```{r summarise_main_dataframe}
summary_bird_migration <- bird_migration |>
                           group_by(Site) |>
                           summarize(
                             iberian_lon = median(iblon, na.rm = TRUE),
                             iberian_lat = median(iblat, na.rm = TRUE),
                             winter_lon = median(winterLongitude, na.rm = TRUE),
                             winter_lat = median(winterLatitude, na.rm = TRUE)
                           )
  
```


```{r merge_tables}
birds <- full_join(summary_bird_migration, bird_breeding_sites)

gt(birds)
```

```{r pivot}
birds_piv <- birds |> 
                  pivot_longer(
                    cols=ends_with(c("_lon", "_lat")),
                    names_to=c('Location_Type', '.value'),
                    names_sep="_") |> arrange(Site, Location_Type)
```

```{r plot_all_points}
ggplot(birds_piv,
       aes(
         x = lon,
         y = lat
       )
  ) +
  geom_point() +
  annotation_borders(
    linewidth=0.2
  )

```

```{r colour_points_type, fig.width=10, fig.height=4.5}
ggplot(birds_piv,
       aes(
         x = lon,
         y = lat,
         color = Location_Type
       )
  ) +
  geom_point() +
  annotation_borders(
    linewidth=0.2
  ) +
  coord_quickmap()

```

```{r colour_points_site, fig.width=10, fig.height=4.5}

ggplot(birds_piv,
       aes(
         x = lon,
         y = lat,
         color = Site
       )
  ) +
  geom_point(size=3) +
  annotation_borders(
    linewidth=0.2
  ) +
  coord_quickmap(
  xlim=c(-25, 100),
  ylim=c(0, 55)
)
```

```{r group}
birds_by_site <- birds_piv |> group_by(Site)
```

```{r join_points_site, fig.width=10, fig.height=4.5}

ggplot(birds_by_site,
       aes(
         x = lon,
         y = lat,
         color = Site,
         shape = Location_Type,
       )
  ) +
  geom_point(size=3) +
  geom_path(aes(group = Site)) +
  annotation_borders(
    linewidth=0.2
  ) +
  coord_quickmap(
  xlim=c(-20, 95),
  ylim=c(0, 60)
  ) + 
  scale_color_brewer(palette = "Set1") + 
  theme_bw()
```