---
title: "Data Frames and Visualisation"
author: "Katy Brown"
date: "2026-07-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/MB1A_Practicals")
library(tidyverse)
library(gt)
library(maps)
library(janitor)
library(lubridate)
```

Download some animal location data from the database https://www.movebank.org/

* On the MoveBank Site, click on "Data" and then "View and Manage Studies". 

* Select "Studies where I can download any data"

* Choose a study from the list. Search for an animal which interests you, if you like. It is best to choose a study with a moderate number of individuals, e.g. between 5 and 15, and 

Rename the file so there are no gaps in the file name.

First I will read the data with the locations of my species into a dataframe.

```{r read_location_data}
locations = read_csv("Dataframes/data/seals.csv") %>%
            clean_names() %>%
            mutate(Date = as_date(timestamp))
```
Look at the dataframe and check that the data looks as expected. Identify the columns containing the latitude and longitude data.

```{r display_location_data}
gt(locations)
```

Isolate the relevant columns. This data is all for one individual.
```{r select_relevant_columns}
loc_small <- locations %>% select(individual_local_identifier, location_long, location_lat, Date)
```

```{r summarise_data}
loc_sum <- loc_small %>%
           group_by(Date, individual_local_identifier) %>%
           summarise(median_long = median(location_long),
                     median_lat = median(location_lat))
```


```{r initiate_plot_1}
ggplot(loc_sum,
       aes(
         x = median_long,
         y = median_lat,
        )
       ) +
  geom_point() +
  annotation_borders(
    linewidth=0.2
  ) +
  coord_quickmap()
```

```{r initiate_plot_1}
ggplot(loc_sum,
       aes(
         x = median_long,
         y = median_lat,
        )
       ) +
  geom_point() +
  annotation_borders(
    linewidth=0.2
  ) +
  coord_quickmap(xlim=c(-98, -92),
                 ylim=c(58, 59.5))
```

```{r join_points_site, fig.width=10, fig.height=4.5}

ggplot(loc_sum,
       aes(
         x = median_long,
         y = median_lat,
         color = individual_local_identifier,
       )
  ) +
  geom_point(size=3) +
  annotation_borders(
    linewidth=0.2
  )  +
  coord_quickmap(xlim=c(-98, -92),
                 ylim=c(58, 59.5)) +
  scale_color_brewer(palette = "Set1") + 
  theme_bw()
```